---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#",
  fig.path = "man/figures/README-",
  out.width = "100%",
  eval = FALSE
)
```

# {yamlsheets}

<!-- badges: start -->
[![Project Status: Concept – Minimal or no implementation has been done yet, or the repository is only intended to be a limited example, demo, or proof-of-concept.](https://www.repostatus.org/badges/latest/concept.svg)](https://www.repostatus.org/#concept)
[![R-CMD-check](https://github.com/matt-dray/yamlsheets/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/matt-dray/yamlsheets/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

## Purpose

Generate spreadsheet publications that follow [best-practice guidance from the UK Government's Analysis Function](https://analysisfunction.civilservice.gov.uk/policy-store/releasing-statistics-in-spreadsheets/), using a [YAML](https://yaml.org/) text file 'blueprint' as the input.

This package is a work-in-progress concept to experiment with new methods for [the {a11ytables} package](https://github.com/co-analysis/a11ytables). It may never be fully-featured or complete.

## Install

Install from GitHub via {remotes}:

```{r}
#| label: install-github
remotes::install_github("matt-dray/yamlsheets")
```

## Quickstart

The basic workflow for {yamlsheets} has three steps:

1. Read a 'blueprint', which is a YAML text file containing all the information you need to create your output spreadsheet.
2. Convert the resulting list object to an {openxlsx2} wbWorkbook-class object, which applies structure and styles.
3. Write the workbook to disk.

In code, that would look something like this:

```{r}
#| label: quickstart-steps
read_blueprint("blueprint.yaml") |>  # pass in your YAML blueprint
  convert_to_workbook() |>  # add structure and styles
  openxlsx2::wb_save("spreadsheet.xlsx")  # write out to an xlsx file
```

You can run this code to see a demo in action:

```{r}
#| label: quickstart-demo
system.file("extdata", "widgets.yaml", package = "yamlsheets") |>  # demo file
  read_blueprint() |> 
  convert_to_workbook() |> 
  openxlsx2::wb_open()  # open a temporary copy
```

## Approach

This section contains some greater depth about how {yamlsheets} works.

### Blueprint files

Information and data to construct a spreadsheet are stored in a particularly-formatted YAML text file, called a 'blueprint'. There's a demo blueprint file provided with the package:

```{r}
#| label: demo-yaml-file
#| eval: true
filepath <- system.file("extdata", "widgets.yaml", package = "yamlsheets")
```

<details><summary>Click to see the raw text content of the blueprint.</summary>

```{r}
#| label: demo-yaml-text
#| eval: true
cat(readLines(filepath), sep = "\n")
```

</details>

<details><summary>Click to read an aside about executing R code in a YAML file.</summary>

You can read a YAML file with `yaml::read_yaml()`. With the argument `eval.expr = TRUE` you can execute R code on-read by pre-pending `!expr` to any lines of your YAML file that you want evaluated as R code. This could, for example, be used to read in data sets stored as CSV files.

The demo blueprint in this package has the line `table: !expr read.csv(system.file("extdata", "widgets.csv", package = "yamlsheets"))`, which reads a demo CSV file that contains the data for Table 1. Let's see what happens when you choose to evaluate that line or not.

```{r}
#| label: demo-yaml-read
#| eval: true
yaml_raw <- yaml::read_yaml(filepath, eval.expr = FALSE)
yaml_evaluated <- yaml::read_yaml(filepath, eval.expr = TRUE)
```

Here you can see the raw, unevaluated R code in the YAML:

```{r}
#| label: demo-yaml-raw
#| eval: true
yaml_raw$table_1$table
```

And here you can see it's been evaluated and the data from the CSV file has been read in:

```{r}
#| label: demo-yaml-executed
#| eval: true
yaml_evaluated$table_1$table
```

</details>

In short, each top-level key represents a sheet (`cover`, etc), with mandatory and arbitrary sub-sections for things like `sheet_title` and `table`. The idea is that these are validated by `read_blueprint()` and interpreted by `convert_to_worbook()` to build each sheet of the workbook.

### Read a blueprint

You can read in the yaml file with `yamlsheets::read_blueprint()`, which is a wrapper around `yaml::read_yaml()` with extra validity checks built in.

Read the blueprint into a list:

```{r}
#| label: demo-read-blueprint
#| eval: true
blueprint <- yamlsheets::read_blueprint(
  yaml = filepath,
  evaluate_r = TRUE,  # default
  print_checks = TRUE  # default
)
```

Note that there are arguments that let us execute R code in the YAML blueprint (`evaluate_r`) and also to print to the console the outcome of some validity checks (`print_checks`). The outcome of each check will appear with a tick or cross to show it has been passed or failed.

Here's a quick preview of the blueprint's structure once it's been read in. You can see it's now a list with one element per sheet.

```{r}
#| label: demo-blueprint-str-1
#| eval: true
str(blueprint, 1)
```

<details><summary>Click to see the full blueprint structure.</summary>

Here's the full structure:

```{r}
#| label: demo-blueprint-str
#| eval: true
str(blueprint)
```

And the full contents in list form:

```{r}
#| label: demo-blueprint-full
#| eval: true
blueprint
```

</details>

### Convert to a workbook

Convert the blueprint list to a wbWorkbook-class object to add spreadsheet structure and style.

```{r}
#| label: demo-workbook
#| eval: true
(workbook <- yamlsheets::convert_to_workbook(blueprint))
```

You can freely manipulate the object with other {openxlsx2} functions if you have additional needs.

Finally, you can open a temporary copy of the workbook:

```{r}
#| label: demo-temp-open
openxlsx2::wb_open(workbook)
```

Or write it to disk:

```{r}
#| label: demo-write
temp_file <- tempfile(fileext = ".xlsx")
openxlsx2::wb_save(workbook, temp_file)
```

## Comparison to {a11ytables}

Improvements in {yamlsheets} compared to {a11ytables} include:

* [{openxlsx2}](https://janmarvin.github.io/openxlsx2/) for the back-end, rather than [{openxlsx}](https://ycphs.github.io/openxlsx/index.html)
* a plain-text 'blueprint' system as a simplified data-input interface (developing [Matt's suggestion](https://github.com/co-analysis/a11ytables/issues/65))
* greater flexibility to provide arbitrary pre-table content
* support for multiple tables per sheet
* use of [{cli}](https://cli.r-lib.org/) and [{fs}](https://fs.r-lib.org/) for improved user interfaces and path-handling

## Related projects

Actively-used packages include:

* [{a11ytables}](https://github.com/co-analysis/a11ytables) for R
* [{rapid.spreadsheets}](https://github.com/RAPID-ONS/rapid.spreadsheets) for R
* ['gptables'](https://github.com/best-practice-and-impact/gptables) for Python

Another experimental project that builds on {a11ytables}:

* [{a11ytables2}](https://github.com/matt-dray/a11ytables2) for R
