---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  eval = FALSE
)
```

# yamlsheets

<!-- badges: start -->
[![Project Status: Concept – Minimal or no implementation has been done yet, or the repository is only intended to be a limited example, demo, or proof-of-concept.](https://www.repostatus.org/badges/latest/concept.svg)](https://www.repostatus.org/#concept)
<!-- badges: end -->

Generate spreadsheet publications that follow [best-practice guidance from the UK government's Analysis Function](https://analysisfunction.civilservice.gov.uk/policy-store/releasing-statistics-in-spreadsheets/), using a [YAML](https://yaml.org/) text file 'blueprint' as the input.

This package is a work-in-progress concept and may never be completed or even reach a useful state.

## Install

Install from R Universe:

```{r}
#| label: install-r-universe
install.packages("yamlsheets", repos = "https://matt-dray.r-universe.dev")
```

Or from GitHub via {remotes}:

```{r}
#| label: install-github
remotes::install_github("matt-dray/yamlsheets")
```

## Approach

Information and data to construct a spreadsheet are stored in a particularly-formatted YAML text file, called a 'blueprint'. There's a demo blueprint provided in the package:

```{r}
#| label: demo-yaml-file
#| eval: true
filepath <- system.file("extdata", "widgets.yaml", package = "yamlsheets")
```

You can view the structure of this demo file by reading manually with `yaml::read_yaml(filepath, eval.expr = TRUE)`.

<details><summary>Click to read an aside about executing R code in a YAML file.</summary>

Note that you can pass data into the YAML file on read by pre-pending `!expr` to R code that will be executed on-read with the argument `eval.expr = TRUE`.

You can choose to do this using the `eval.expr` argument in `yaml::read_yaml()`:

```{r}
#| label: demo-yaml-read
#| eval: true
yaml_raw <- yaml::read_yaml(filepath, eval.expr = FALSE)
yaml_evaluated <- yaml::read_yaml(filepath, eval.expr = TRUE)
```

Here you can see the raw, unevaluated R code in the YAML:

```{r}
#| label: demo-yaml-raw
#| eval: true
yaml_raw$table_1$table
```

And here you can see it's been evaluated and the data from the CSV file has been read in:

```{r}
#| label: demo-yaml-executed
#| eval: true
yaml_evaluated$table_1$table
```

</details>

Read the blueprint  into a list:

```{r}
#| label: demo-read-blueprint
#| eval: true
blueprint <- yamlsheets::read_blueprint(filepath)
str(blueprint, 1)
```

<details><summary>Click to see the full blueprint structure.</summary>

```{r}
#| label: demo-blueprint-full
#| eval: true
str(blueprint)
```

</details>

Convert the blueprint to a wbWorkbook-class object:

```{r}
#| label: demo-workbook
#| eval: true
(workbook <- yamlsheets::convert_to_workbook(blueprint))
```

You can freely manipulate the `workbook` object with other {openxlsx2} functions.

You can open a temporary copy of the workbook:

```{r}
#| label: demo-temp-open
openxlsx2::wb_open(workbook)
```

Or write it to disk:

```{r}
#| label: demo-write
temp_file <- tempfile(fileext = ".xlsx")
openxlsx2::wb_save(workbook, temp_file)
```

## Related projects

* [{a11ytables}](https://github.com/co-analysis/a11ytables) is an R package.
* [{rapid.spreadsheets}](https://github.com/RAPID-ONS/rapid.spreadsheets) is an R package.
* ['gptables'](https://github.com/best-practice-and-impact/gptables) is a Python package.
